Final paper: Towards Proving Termination of \texttt{zelda-mosaic} Algorithms under Conditions on the Input

### 2021-05-03 paper/paper.tex
paper: bring over justin's work & copy-edit

\matlab\ supports a number of operations that can be used to manipulate arrays,
but only a subset are used in our program. To illustrate the operations used, we
begin with a visualization of a three-dimensional array \(E\) with dimensions
\(3 \times 4 \times 2\) below and discuss how it is used in \matlab.

@@

@@

\section{Old / Superceded}

\subsection{Data Structures and Functions}

Our program is used for image processing; thus, many computations are made on
images, which are represented as arrays in \matlab\@. Additionally, we deal with
matrices of images, so we need to model multi-dimensional matrices. Since
multi-dimensional matrices can have arbitrarily many dimensions, we decided the
best abstraction for this in Coq would be built on top of the \textsf{Vector}
type which was originally implemented in the \colorlib\
library~\cite{BLANQUI_2011} and eventually ported to the Coq standard library.
In Coq, \textsf{Vector} is a dependent type~\cite{Bove2009,Thorsten_2010}; the
number of elements in a \textsf{Vector} is part of its type, written
\texttt{Vector.t A n} for a vector containing \(n\) elements of type \(A\).
Since \textsf{Vector}s are polymorphic, we noted that a \textsf{Vector} could
contain \textsf{Vector}s provided they are all the same type (which includes
length, as noted). That is, we can stack \textsf{Vector}s to form, \emph{e.g.},
\texttt{Vector.t (Vector.t A m) n}. From there, we defined a type-constructor
\textsf{Matrix} taking a list of dimensions, which represents a \textsf{Vector}
of \textsf{Vector}s. It can be used as the type of an arbitrary
\(n\)-dimensional matrix containing elements of type \(A\).

Due to the implementation of \textsf{Vector} in the Coq standard library, we
needed to greatly extend the functionality of our \textsf{Matrix} type. To that
end, we have begun implementing \textsf{Matrix} functions such as element
access, which is non-trivial for matrices with an arbitrary number of
dimensions\footnote{In particular because of the dependent types; see also
\url{https://stackoverflow.com/q/66863226/4400820}}. Additionally, the existing
implementation of element access in Coq's \textsf{Vector} type uses the
unfamiliar \textsf{Fin} type, an abstraction for carrying proofs that a
particular number belongs to finite range of numbers. In order to access an
element of a \textsf{Vector}, one must supply such a proof; for our purposes, we
use \texttt{Fin.of\_nat}, which gives either such a proof or a witness that the
number is outside the range. This behavior is captured by the \textsf{sumbool}
built-in, which required further research on our behalf. Given the number of
ways that matrices can be indexed in \matlab\ and the complexity introduced by
Coq's implementation of \textsf{Vector}s, we see this as the main bottleneck for
our progress. We still need to implement many \textsf{Matrix} operations, such
as accesses over ranges and over an entire dimension.

\subsection{Binding Operations and Semantics}

While we have not yet bound any of our functions or data types to the statements
or expressions in \S~\ref{S:mmatlab}, we have naturally been thinking about them
throughout the whole process. The functions we write are intended to implement
the evaluation of the corresponding expressions. In the style of
\texttt{Imp}~\cite{Pierce:SF1}, we intend to write a deterministic expression
evaluation function; we will then provide semantics for program execution.
Execution cannot be defined as a function in Coq because it may be
non-terminating. We have had to consider what operations are be allowed and
whether checks for undesired behavior (\emph{e.g.}, dynamic type mismatches)
should be implemented in the underlying functions or in a semantic check.

\section{Goals and Timeline}

As we neglected to include them in our original proposal, we now propose the
following goals, taking into consideration the progress that we have made so far
with the project:

\begin{enumerate}

    \item \textbf{Minimum Viable Product}: Successful modelling of a majority of
        the desired features corresponding to matrix operations in \mmatlab\@.
        Given the lack of multi-dimensional matrix libraries in Coq and the
        necessary additions of several features in the Coq standard library,
        this alone would be a notable achievement for us.

    \item \textbf{Desired Goal}: Successful binding of our matrix operations to
        \mmatlab\ statements and expressions, including the ability to evaluate
        programs written in \mmatlab\@. A proof of termination, or sketch of
        one, for our V1 algorithm is provided.

    \item \textbf{Stretch Goal}: All of the above, and the addition of all
        operations used in V2 and V3 algorithms (most notably, for loops that
        increment across a range at varying intervals). Proofs that use
        these features.

\end{enumerate}

Given what we have encountered so far, we propose the following modified
timeline for project completion.

\begin{itemize}
    \item \textbf{03/31/21}: Progress Report Due
    \item \textbf{04/09/21}: Significant progress made on most matrix operations
        and data structures.
    \item \textbf{04/16/21}: Binding \mmatlab\ syntax to operations written
        in Coq; definition of semantics and program evaluator; formal proof
        specification written.
    \item \textbf{04/24/21}: Proof(s) should be written or scope adjusted
        accordingly. All logic should be implemented; edits for readability
        should be the main changes henceforth.
    \item \textbf{04/26/21}: Final Proof Due
    \item \textbf{05/03/21}: Final Paper Due
    \item \textbf{05/05/21}: Final Presentation Delivered
\end{itemize}

We provide the following ``diff'' between the scope of our original project
proposal and our new proposal:
\begin{itemize}
    \item Heightened emphasis on implementation of matrix operations, since they
        are proving to be non-trivial.
    \item Addition of goal tiers.
    \item Discussion on Hoare logic.
\end{itemize}

>>>>

\matlab\ supports a number of operations that can be used to manipulate arrays,
but only a subset are used in our program. To illustrate the operations used, we
begin with the following visualization of a three-dimensional array \(E\) with
dimensions \(3 \times 4 \times 2\) and discuss how it is used in \matlab.

@@

We made an effort to specify each of the statements and expressions according to
Backus-Naur form, considering the syntax of which each statement or expression
was composed and instances of overloading. Our next goal was to model the state
of our program's core matrix data structure so that we could bind syntax to
semantics. For reasons that we will soon delve into, we did not progress further
than that step, so our discussion on syntax and semantics ends here. However,
had we gotten to the point of tying it all together, it would have looked very
similar to the \textit{Simple Imperative Programs} chapter in \textit{Logical
Foundations}~\cite{Pierce:SF1}.

\subsection{Blocked by Built-ins}

Thinking that there was no need to re-invent the wheel, we began our journey in
defining matrix behavior with an exploration into a standard Coq library. Along
the way, we discovered a number of interesting implementations and proof
paradigms. Ultimately, we concluded that these would be too difficult to work
with and that we were better off developing our own implementation. Here we
discuss the structures from the Coq library that we considered using and the
limitations that led us away from them.

\subsubsection{Vector}

The natural candidate in the Coq library to use for representing
multi-dimensional matrices was the \textsf{Vector} type, which was originally
implemented in the \colorlib\ library~\cite{BLANQUI_2011} and later ported to
the Coq standard library. We thought that this would be a good abstraction since
multi-dimensional matrices can have arbitrarily many dimensions and
\textsf{Vector}s are polymorphic, allowing them to contain other
\textsf{Vector}s (provided that all contained vectors are of the same type). To
aid this discussion, the definition of Coq's \textsf{Vector} is stated below:

\begin{verbatim}
Inductive t A : nat -> Type :=
| nil : t A 0
| cons : forall (h:A) (n:nat),
           t A n -> t A (S n).
\end{verbatim}

The key thing to make note of is that \textsf{Vector} is a dependent
type~\cite{Bove2009,Thorsten_2010}; the number of elements in a \textsf{Vector}
is part of its type, written \texttt{Vector.t A n} for a vector containing \(n\)
elements of type \(A\)\footnote{The Coq programmer works with dependent types
all the time, but may not realize it: hypotheses whose types depend on
quantified variables (\textit{e.g.}, the hypothesis \(H: n \le 0\)) are
dependently typed as well.}. Since the length of a particular \textsf{Vector} is
encoded in its type, it is possible to enforce the requirement that all elements
in a particular dimension have the same length. From here, we devised a scheme
to stack \textsf{Vector}s to form, \emph{e.g.}, \texttt{Vector.t (Vector.t A m)
n} such that this we could create matrices of arbitrarily many dimensions, each
of which has a fixed number of elements. To support this, we defined a
type-constructor \textsf{matrix} taking a list of dimensions, which represents a
\textsf{Vector} of \textsf{Vector}s. It can be used as the type of an arbitrary
\(n\)-dimensional matrix containing elements of type \(A\). This definition is
as follows:

\begin{verbatim}
Fixpoint matrix
        (A: Type) (dims: list nat) :=
  match dims with
  | [] => A
  | head::tail =>
        Vector.t (matrix A tail) head
  end.

\end{verbatim}

This definition worked out nicely at first, and we were able to implement
behavior such as the linearization and element access of a matrix. Although
possible, this task was non-trivial due to the arbitrary dimensionality of our
matrices and due to the proof burden imposed on functions over dependent
types~\cite{SO_2021_1}. Additionally, the existing implementation of element
access in Coq's \textsf{Vector} type uses the unfamiliar \textsf{Fin} type, an
abstraction for carrying proofs that a particular number belongs to finite range
of numbers. In order to access an element of a \textsf{Vector}, one must supply
such a proof; for our purposes, we use \texttt{Fin.of\_nat}, which gives either
such a proof or a witness that the number is outside the range. The complexity
of using the Coq built-ins was quickly growing, and we could deal with it to a
point but soon ran into trouble while implementing ranged expressions.

\subsubsection{Out of Options}

The problem we experienced while implementing ranged expressions is complex, so
here we provide an attempt to explain the essence of the problem. The full
details are available on StackOverflow, where we asked for
assistance~\cite{SO_2021_2}. The general flavor of the problem is that when
accessing indices over a range in a matrix, it is impossible to know \textit{a
priori} the shape of the returned matrix, which in turn informs the returned
matrix's type (since it is dependently typed based on its dimensionality). This
computation can be made dynamically for valid ranges, but we run into issues
when handling degenerate cases---for instance, a range containing values that
are out of bounds or a list of indices containing the wrong number of indices
for a particular matrix. In these situations, it is desirable to not return a
matrix type at all, since the behavior is undefined and will not result in a
legitimate matrix. The crux of the problem is that with the dependent types, it
was difficult to specify what dimensions an invalid matrix should have, and the
general strategy of using \textsf{Option}'s \textsf{Some} and \textsf{None}
wrappers also did not work because it requires type specification (which is what
we were having difficulty doing in the first place). Presented with this
dilemma, we heeded the advice of a commenter on our StackOverflow post, which
was to eschew dependent types altogether and build our own implementation of
matrices. The results of doing so encompass the main body of our work, and are
described in detail in \S~\ref{S:results}.

@@
